Measurement Bias from Address Aliasing

### 2014-11-19 paper.tex
Improved table and figure layout.

\begin{table}
  \caption{Experimental setups used.\label{tab:machines}}{
  \begin{tabular}{|l|l|}
    \hline
Core          & Intel\textsuperscript{\textregistered{}} Core\texttrademark{}2 Duo P8600 (laptop) \\
              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0), GCC 4.8.2-19ubuntu1 \\ \hline
Nehalem       & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-950 \\
              & 64-bit Ubuntu 12.04 LTS (Linux 3.2.0), GCC 4.8.2-19ubuntu1 \\ \hline
% Sandy Bridge  & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-2670QM (laptop) \\
%               & 64-bit Ubuntu 11.04 (Linux 3.0.0) \\ \hline
Ivy Bridge    & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i5-3470 \\
              & 64-bit Ubuntu 12.04 LTS (Linux 3.8.0), GCC 4.8.2-19ubuntu1 \\ \hline
Haswell       & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-4770K \\
              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0), GCC 4.8.2-19ubuntu1 \\ \hline
  \end{tabular}}
\end{table}

@@

% See /analysis for results showing this does not alias
\begin{figure}
  \begin{lstlisting}[frame=single, xleftmargin=.1\textwidth, xrightmargin=.1\textwidth]
#define ALIAS(a, b) \
    ((((long)&a) & 0xfff) == (((long)&b) & 0xfff))
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    if (ALIAS(inc, i) || ALIAS(g, i))
        return main();
    for (; g < 65536; g++) {
        i += inc;
        j += inc;
        k += inc;
    }
    return 0;
}
  \end{lstlisting}
  \caption{Modified microkernel that can dynamically detect aliasing case, and avoid it by pushing another stack frame.}
  \label{lst:loopfixed}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \lstinputlisting[
    language=C,
    frame=single,
    xleftmargin=.1\textwidth,
    xrightmargin=.1\textwidth
  ]{bin/convolution-kernel.c}
  \caption{Basic implementation of convolution with a fixed kernel, ignoring endpoints for simplicity. This program is highly sensitive to aliasing between input and output arrays.}
  \label{lst:conv}
\end{figure}

@@

\pgfplotstableread{bin/conv-default-o2-haswell.estimate.dat}{\convtabletwohaswell}
\pgfplotstableread{bin/conv-default-o3-haswell.estimate.dat}{\convtablethreehaswell}
\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        title=Haswell -O2,
        font=\footnotesize,
        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/1.9, % Make it fit to text width side by side
        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwohaswell ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwohaswell ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=Haswell -O3,
        font=\footnotesize,
        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/1.9,
        skip coords between index={20}{32}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethreehaswell ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethreehaswell ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}. Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Showing results for optimization levels O2 and O3 compiled with GCC, input size $n=2^{20}$.}
\end{figure}

@@

We look at three different libraries available through the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$ for \texttt{double} sized numbers.
In this equation,  $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix.
We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
Denote the matrix size as $M \times N$, making $\boldsymbol{x}$ of length $N$ and $\boldsymbol{y}$ of length $M$.
In code, the function is invoked with the following parameters:
\begin{lstlisting}
cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
\end{lstlisting}

@@

\pgfplotstableread{bin/libblas.dat}{\libblastable}
\pgfplotstableread{bin/libatlas.dat}{\libatlastable}
\pgfplotstableread{bin/libopenblas.dat}{\libopenblastable}
\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        title=libblas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8, % Make it fit to text width side by side
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libblastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libatlas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8,
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libatlastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libatlastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libopenblas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8,
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libopenblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libopenblastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:blas}Performance of \texttt{cblas\_dgemv} for different BLAS packages, measuring cycles executed and address alias events for varying relative address offset between matrix $A$ and vector $\boldsymbol{y}$. Each sample point represents an increment of 16 bytes, or 0x10.}
\end{figure}

>>>>

\begin{table}
  \caption{Experimental setups used.\label{tab:machines}}{
  \begin{tabular}{|l|l|}
    \hline
Core          & Intel\textsuperscript{\textregistered{}} Core\texttrademark{}2 Duo P8600 (laptop) \\
              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0),\\
              & GCC 4.8.2-19ubuntu1 \\ \hline
Nehalem       & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-950 \\
              & 64-bit Ubuntu 12.04 LTS (Linux 3.2.0),\\
              & GCC 4.8.2-19ubuntu1 \\ \hline
% Sandy Bridge  & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-2670QM (laptop) \\
%               & 64-bit Ubuntu 11.04 (Linux 3.0.0) \\ \hline
Ivy Bridge    & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i5-3470 \\
              & 64-bit Ubuntu 12.04 LTS (Linux 3.8.0),\\
              & GCC 4.8.2-19ubuntu1 \\ \hline
Haswell       & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-4770K \\
              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0),\\
              & GCC 4.8.2-19ubuntu1 \\ \hline
  \end{tabular}}
\end{table}

@@

% See /analysis for results showing this does not alias
\begin{figure}
  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
#define ALIAS(a, b) \
    ((((long)&a) & 0xfff) == (((long)&b) & 0xfff))
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    if (ALIAS(inc, i) || ALIAS(g, i))
        return main();
    for (; g < 65536; g++) {
        i += inc;
        j += inc;
        k += inc;
    }
    return 0;
}
  \end{lstlisting}
  \caption{Modified microkernel that can dynamically detect aliasing case, and avoid it by pushing another stack frame.}
  \label{lst:loopfixed}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \lstinputlisting[
    language=C,
    frame=single,
    xleftmargin=.01\textwidth,
    xrightmargin=.01\textwidth
  ]{bin/convolution-kernel.c}
  \caption{Basic implementation of convolution with a fixed kernel, ignoring endpoints for simplicity. This program is highly sensitive to aliasing between input and output arrays.}
  \label{lst:conv}
\end{figure}

@@

\pgfplotstableread{bin/conv-default-o2-haswell.estimate.dat}{\convtabletwohaswell}
\pgfplotstableread{bin/conv-default-o3-haswell.estimate.dat}{\convtablethreehaswell}
\begin{figure}
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        title=Haswell -O2,
        font=\footnotesize,
        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/1.9, % Make it fit to text width side by side
        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwohaswell ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwohaswell ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=Haswell -O3,
        font=\footnotesize,
        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/1.9,
        skip coords between index={20}{32}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethreehaswell ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethreehaswell ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}. Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Showing results for optimization levels O2 and O3 compiled with GCC, input size $n=2^{20}$.}
\end{figure}

@@

We look at three different libraries available through the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$ for \texttt{double} sized numbers.
In this equation,  $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix.
We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
Denote the matrix size as $M \times N$, making $\boldsymbol{x}$ of length $N$ and $\boldsymbol{y}$ of length $M$.
In code, the function is invoked with the following parameters:
\begin{lstlisting}[breaklines=true]
cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
\end{lstlisting}

@@

\pgfplotstableread{bin/libblas.dat}{\libblastable}
\pgfplotstableread{bin/libatlas.dat}{\libatlastable}
\pgfplotstableread{bin/libopenblas.dat}{\libopenblastable}
\begin{figure}
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        title=libblas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8, % Make it fit to text width side by side
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libblastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libatlas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8,
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libatlastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libatlastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libopenblas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8,
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libopenblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libopenblastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:blas}Performance of \texttt{cblas\_dgemv} for different BLAS packages, measuring cycles executed and address alias events for varying relative address offset between matrix $A$ and vector $\boldsymbol{y}$. Each sample point represents an increment of 16 bytes, or 0x10.}
\end{figure}
